Machine Learning Methods for Discourse Segmentation of Communications in E-Mail Based Behavioral Interventions

### 2018-07-30 paper-amia-summit-2018/manuscript.tex
update paper

\noindent{\bf Abstract}
\textit{Communication science approaches to developing effective behavior interventions, such as motivational interviewing (MI), are limited by the traditionally manual qualitative coding of communication exchanges, which is a very resource-intensive and time-consuming process. This study focuses on the analysis of e-Coaching sessions, behavior interventions that are delivered via email and grounded in the principles of MI. A critical step towards automated annotation of e-Coaching communication exchanges is segmentation of emails into textual fragments that correspond to MI behaviors. In this work, we formulate this task as a classification problem and propose lexical, punctuation and part-of-speech features to address it. We experimented both with traditional machine learning method, conditional random fields (CRF) and deep learning methods, such as multilayer perceptrons (MLP), bidirectional recurrent neural networks (BRNN) and convolutional recurrent neural networks (CRNN). Results indicate that CRNN outperformed CRF, MLP and BRNN achieving 0.989 macro F1-score overall and 0.825 macro F1-score for detecting new segment.}

\section*{Introduction}
The emergence of e-Health technologies opened up new ways to deliver a variety of behavioral interventions to any demographic group of patients in any geographical location. Motivational interviewing (MI), an evidence-based communication technique to increase intrinsic motivation and self-efficacy for behavior change\cite{miller2012motivational,miller2009ten,miller2009toward}, is one type of these interventions. MI sessions are generally aimed at eliciting ``change talk'', or statements of intrinsic motivation about patients' own desire, ability, reasons, need for and commitment to behavior change, which have been established by previous research\cite{apodaca2009mechanisms} as a reliable mediator of health behavior change. However, communication science approaches to understanding the efficacy of MI are inherently limited by traditional qualitative coding methods.

Qualitative coding of motivational interviews with pre-defined codes, has been traditionally performed manually by trained annotators which is a tedious and resource-intensive process that involves several iterations of reading, comprehension and interpretation of interview transcripts. Rapidly developing computational technologies, specifically, machine learning methods, offer a unique opportunity to accelerate this process. In particular, machine learning methods have been successfully applied to a variety of analytical tasks involving textual data, such as classification\cite{nigam2000text} and sentiment analysis\cite{wang2012baselines}. In our previous work, we examined the utility of machine learning methods for automated annotation \cite{hasan2016study,kotov2015interpretable} and analysis \cite{hasan2018predicting} of in-person MI sessions. Specifically, we demonstrated that machine learning methods can be utilized for annotation of MI transcripts according to a simple communication code scheme with the accuracy comparable to human coders\cite{hasan2016study}. Experimental data utilized in these studies, however, were prepared by transcribing audio conversations, which were clearly segmented into utterances by a counselor, a patient and a caregiver.

In this study, we focus on the analysis of e-Coaching sessions, behavior interventions that are delivered via email and grounded in the principles of motivational interviewing. Specifically, the e-Coaches involved in this study used emails to communicate motivation-enhancing messages that encourage healthy eating among GenY adolescents. e-Coaching data is comprised of email responses, which are free-text documents, unlike more traditional dyadic clinical interviews that are naturally segmented into utterances due to their conversational nature.

The unstructured nature of e-Coaching exchanges poses a unique set of challenges for their qualitative analysis. Usually, email segments are automatically code with the Minority Youth-Sequential Coding of Process Exchanges (MYSCOPE \cite{carcone2013provider}), a qualitative coding scheme to characterize client-coach communication during email exchange. A significant barrier to fully automating the behavior coding process of e-Coaching emails is their segmentation into textual fragments that correspond to distinct communication behaviors. Automating this task is a unique and challenging problem due to the following major reasons:

@@

The goal of this research study is to assess the applicability of machine learning methods for automated segmentation of e-Coaching emails into textual fragments corresponding to individual behaviors, which is the first step of the coding process of e-Coaching communications. In particular, we introduced lexical, punctuation and  part-of-speech (POS) features and experimented with both traditional supervised machine learning method, linear-chain conditional random fields (CRF) and deep learning methods, such as multilayer perceptrons (MLP), bidirectional recurrent neural networks (BRNN) and convolutional recurrent neural networks (CRNN), to find the best performing method and feature combination.

Relevant previous work in the biomedical domain primarily focused on segmentation of text into sections and headers\cite{apostolova2009automatic,denny2009evaluation,tepper2012statistical,cho2002text} or sentence boundary detection\cite{griffis2016quantitative,kreuzthaler2015detection,treviso2017sentence,fraser2015sentence}. Apostolova et al.\cite{apostolova2009automatic} applied SVM along with word-vector cosine similarity metric combined with several heuristics to segment clinical reports into sections, such as demographics, history, procedure, finding and impression. After identification of each line in the document, Tepper et al. \cite{tepper2012statistical} trained Maximum Entropy models for section classification. Denny et al.\cite{denny2009evaluation} proposed a SecTag algorithm, which combined natural language processing techniques, terminology-based rules and Naive Bayes classifier to identify the sections and headers that achieved 99\% recall with 95.6\% precision. On the other hand, SVM based on prosodic and part of speech features \cite{kreuzthaler2015detection} and recurrent convolutional neural networks using word embeddings \cite{griffis2016quantitative} were utilized for detecting sentence boundaries. Liu et al. have shown that a linear chain CRF performed better than the hidden markov model and Maxent on the NIST sentence boundary detection \cite{liu2005using}. Recently, Treviso et al. proposed a recurrent convolutional neural network for automatic identification of sentence boundary in a neuropsychological text in Portuguese language \cite{treviso2017sentence}. Their proposed model combined prosodic and part of speech features with pre-trained GloVe word vectors \cite{pennington2014glove}, which achieved 0.74 F1-score in sentence boundary detection in the Brazilian constitution dataset.

In this study, we adopt the model proposed by Treviso et al. \cite{treviso2017sentence} and apply it differently from the aforementioned studies. The main difference is that our study focuses on segmentation of e-Coaching emails as distinct from determining the sentence boundary. Another important difference is that punctuation features were utilized in combination with word embeddings and part-of-speech features. Furthermore, we train our model with pre-trained word2vec word vectors \cite{mikolov2013distributed}, instead of using GloVe word vectors. Segmentation of e-Coaching emails is also different from a traditional shallow discourse analysis of conversations \cite{galley2003discourse} in that the focus is on segmentation, rather than on determining the types of transitions between the utterances or assigning utterances to speakers.

Recently, an online clinical intervention called MENU GenY \cite{alexander2017motivations} (Making Effective Nutrition Choices for Generation Y) was proposed and evaluated. MENU GenY is a technology-based public health intervention that relies on personalized e-coaching to encourage increased fruit and vegetable intake among young adults, aged 21-30. The goal of MENU GenY was to develop a better coding dictionary among GenY to improve eating habits. However, segmentation of clinical exchange in the context of electronically delivered interventions, in particular, segmentation of clinical interaction text into groups of MI behaviors, is still performed manually, which slows down qualitative analysis of these interventions. This study proposes several machine learning methods to address this problem and the authors are unaware of any other work that focused on the same problem.

@@

The experimental dataset for this work was constructed from 49 e-coaching sessions, which include a total of 3,138 segmented and annotated MI behaviors. Each session represents an MI intervention delivered via email. We formulate the segmentation task as a binary classification problem where each word or punctuation mark annotated with a label 1 or 0, to indicate whether it precedes a new segment or not. In total, we obtained 95,777 words and 7,140 punctuation marks, which include 3,138 ``new segment'' and 99,779 ``same segment'' instances. In this study, we experimented with traditional machine learning method, CRF and deep learning methods, such as MLP, BRNN and CRNN. For MLP models, samples were created with $2l$ words/punctuations at each step in a word sequence, where a sample contains next $l$ words/punctuations and prior $l$ words/punctuations including current word/punctuation. Each sample is classified into either ``new segment'' or ``same segment'' based on whether the current word/punctuation precedes a new segment or not. For CRF, BRNN and CRNN models, an email was taken as input sequence, such that POS tags and word embedding of each word or punctuation mark were used as input and binary labels (1 or 0) corresponding to ``new segment'' and ``same segment'' classification decision were considered as the output of the model at each step. In the gold standard, words or punctuations within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1.

@@

We utilized three types of features in conjunction with CRF, MLP, BRNN and CRNN methods: word embeddings or lexical features, punctuation and POS features. Since POS tags have been shown to be effective semantic abstractions of individual words, we used POS features for our experiment \cite{liu2005using,treviso2017sentence}. To extract our POS features, we tag the e-Coaching emails using the NLTK POS tagger \cite{bird2009natural}. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;'\} between a pair of words, are also employed as a feature since punctuation marks designate the boundary of a sentence, clause and phrase and often also correspond to a segment boundary. For natural language processing (NLP) tasks, inputs are received as a text where words can be considered as the basic processing unit. Therefore, it is important to represent a word in such a way that it carries all relevant information. Word embeddings are one of this representation where each word represented as a real-valued vector in a high dimensional vector space. The distributed representation of word allows capturing semantic, syntactic and morphological information from large unannotated corpora  \cite{pennington2014glove, mikolov2013distributed}. Since pre-trained word2vec word vectors provide the best results, in our experiment, we utilized pre-trained word2vec word vectors for words. Trained word vectors were also utilized when rare words or punctuation marks are not available in the pre-trained word2vec word vectors. Figure~\ref{fig:embedding-dimension} illustrates the performance of CRNN model on e-Coaching email segmentation by varying the dimension of pre-trained and trained word embeddings with GloVe and word2vec models. It was observed that best performance achieved with pre-trained 300 dimensional word2Vec word vectors when three types of features are used together. Therefore, we report our results with word2vec 300 dimensional word vectors for all deep learning models used in this study. CRF utilized lexical features which represent each word in an e-Coaching email.

@@

In the present study, we experimented with four different classifiers, including one traditional machine learning method, CRF and three deep learning models. Since deep learning provides a flexible architecture for constructing and synthesizing complex models, we take the advantage of this flexibility to test MLP, BRNN and CRNN models for the task of segmentation of e-Coaching emails.

\textbf{Conditional Random Fields}: CRF have been widely used in various NLP tasks including part-of-speech tagging and segmentation tasks \cite{lafferty2001conditional, hirohata2008identifying}. Unlike a maximum entropy markov model which uses per-state exponential models for the conditional probabilities of next states given the current state, a CRF is a discriminative model of the conditional distribution of the entire sequence of labels given the observation sequence. A traditional linear-chain CRF model is defined as a conditional probability distribution $p(y|x)$ for output and input sequences, $y$ and $x$, by the Eq. \ref{Eq:crf}

@@

Here $Z_x$ is a normalization factor over all possible labelings of $x$, and $f_k(y_{t-1}, y_t, x, t)$ is a feature function, and $\lambda_k$ is a learned weight associated with feature $f_k$. The optimal output sequence $y^*$ for an input sequence $x$, $y^* = {arg\,max}_y p(y|x)$, is obtained efficiently by the Viterbi algorithm. In our experiments, the following features were utilized with CRF models: i) current word/punctuation ii) next and previous 3 words/punctuations iii) whether the word/punctuation is special characters (';', '?', '.', ',', '!', ':', etc) or not iv) whether the word is title word or not? (e.g. ``The'' is a title word but ``the'' is not) v) POS tags

\textbf{Multilayer Perceptrons}: MLP is a feed-forward artificial neural networks which map set of input onto a set of appropriate output. This study utilized MLP with a nonlinear activation function (relu) and one hidden layer consisting of 128 hidden units. In order to prevent over-fitting, we implement dropout on fully connected layers, which randomly hide some neurons during the training phase \cite{srivastava2014dropout}. Dropout is also applied in the fully connected layers in BRNN and CRNN models. For a given sample with $2l$ words/punctuations, the first layer input of MLP was created by the summation of first $l$ word vectors and last $l$ word vectors. Figure~\ref{fig:length-mlp} demostrates the performance of MLP model on e-Coaching email segmentation by varying value of $l$. Based on the results in Figure~\ref{fig:length-mlp}, MLP results were reported with $l = 2$, since best performance of MLP models were observed when $l = 2$.

@@

\textbf{Bidirectional Recurrent Neural Networks}: BRNN is a neural network designed to capture sequential patterns by considering both past and future inputs as well as complex relationships between input feature and output labels. The output of BRNN layer is computed as the summation of the forward RNN output with backward RNN output. Gated Recurrent Units (GRU) \cite{chung2014empirical} are utilized as RNN that are capable of handling variable size input sequence and have an internal memory that can be reset. Figure~\ref{fig:crnn} will represent an architecture of BRNN if convolution layer is removed.

\textbf{Convolutional Recurrent Neural Networks}: CRNN is a deep neural networks architecture, shown in Figure~\ref{fig:crnn}, consisting of 5 layers: 1) input layer 2) embedding layer 3) convolution layer with max pooling 4) BRNN layer 5) fully connected layer with dropout and sigmoid output.

@@

A sequence of $m$ words is fed into the input layer which provides a $m \times n_e$ matrix after fetching the pre-trained word vectors. When POS tags are used in combination with word embedding, 10-dimensional POS vectors are concatenated with 300-dimensional word vectors to obtain new vectors $n_e = n_w + n_p$ of size 310. The 1D convolutional layer generates $m - h_c + 1$ new features with $h_{c.n_e}$, a filter of $h_c$ words and zero-padding on both sides of the text. A pooling operation was performed over time to find the most significant features in a region. After applying $n_f=100$ different filters on convolutional layer, $m \times n_f$ inputs are generated and fed into bidirectional recurrent neural networks, which contain 200 hidden units. Then, dropout is used to randomly drop 50\% of the neurons and passes them through a fully connected layer, where a sigmoid operation is calculated, giving us the probability of whether the word or punctuation precedes a new segment or not. We experimentally determined that the best performance is achieved with 5 folds corss-validation when filter length in convolution layer is 7, pool size is 3, batch size is 32, adam\cite{kingma2014adam} is used for optimization and the early stopping strategy is used with dropout to prevent over-fitting. To reuse of our source codes for other research studies, source codes of all experiments will be publicly available at https://github.com/teanalab/xxxxx upon acceptance of this paper for publication.

\subsection*{\textit{Evaluation metrics}}
We report standard metrics for experiments (precision, recall and F1-measure) to evaluate the performance of binary classifiers\cite{aas1999text}. However, accuracy is not reported as a performance metric because accuracy is highly sensitive to the prior class probabilities and does not fully describe the actual difficulty of the decision problem for an unbalanced dataset. The results are reported based on 5 folds cross-validation and weighted macro-averaging over the folds. We also estimate the area under the receiving operating characteristics (ROC) curve\cite{kumar2011receiver} (AUC) metric due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets \cite{hu2015kernelized}.

@@

\begin{table}[ht]
\centering
\caption{Performance of CRF, MLP, BRNN and CRNN methods for identification of ``new segment'' class as well as weighted average over ``new segment'' and ``same segment'' classes when word embeddings or lexical features are used. The highest value for each performance metric
is highlighted in bold.}
\label{tab:result_base}
  \begin{tabular}{|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{new segment}} & \multicolumn{3}{|c|}{\textbf{weighted macro average}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} \\ \hline    
 CRF & 0.782 & 0.691 & 0.733 & 0.983 & 0.984 & 0.984 \\ \hline
 MLP & \textbf{0.836} & 0.593 & 0.694 & 0.982 & 0.983 & 0.982 \\ \hline
 BRNN & 0.606 & 0.680 & 0.641 & 0.977 & 0.976 & 0.976 \\ \hline
 CRNN & 0.775 & \textbf{0.797} & \textbf{0.785} & \textbf{0.986} & \textbf{0.986} & \textbf{0.986} \\ \hline
  \end{tabular}
\end{table}

@@

\begin{table}[ht]
\centering
\caption{Performance of CRF, MLP, BRNN and CRNN methods for identification of ``new segment'' class as well as weighted average over ``new segment'' and ``same segment'' classes when all features are used together. The highest value for each performance metric
is highlighted in bold.}
\label{tab:result_weighted_avg}
 \begin{tabular}{|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{new segment}} & \multicolumn{3}{|c|}{\textbf{weighted macro average}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} \\ \hline    
 CRF & 0.813 & 0.772 & 0.792 & 0.988 & 0.988 & 0.988 \\ \hline
 MLP & \textbf{0.814} & 0.732 & 0.771 & 0.987 & 0.987 & 0.987 \\ \hline
 BRNN & 0.683 & 0.820 & 0.745 & 0.985 & 0.983 & 0.984 \\ \hline
 CRNN & 0.789 & \textbf{0.864} & \textbf{0.825} & \textbf{0.990} & \textbf{0.989} & \textbf{0.989} \\ \hline
  \end{tabular}
\end{table}

>>>>

\noindent{\bf Abstract}
\textit{Communication science approaches to developing effective behavior interventions, such as motivational interviewing (MI), are limited by the traditionally manual qualitative coding of communication exchanges, which is a very resource-intensive and time-consuming process. This study focuses on the analysis of e-Coaching sessions, behavior interventions that are delivered via email and grounded in the principles of MI. A critical step towards automated annotation of e-Coaching communication exchanges is segmentation of emails into textual fragments that correspond to MI behaviors. In this work, we formulate this task as a classification problem and propose word embedding, punctuation and part-of-speech features to address it. We experimented both with traditional machine learning method, conditional random fields (CRF) and deep learning methods, such as multilayer perceptrons (MLP), bidirectional recurrent neural networks (BRNN) and convolutional recurrent neural networks (CRNN). Results indicate that CRNN outperformed CRF, MLP and BRNN achieving 0.989 macro F1-score overall and 0.825 macro F1-score for detecting new segment.}

\section*{Introduction}
The emergence of e-Health technologies opened up new ways to deliver a variety of behavioral interventions to any demographic group of patients in any geographical location. Motivational interviewing (MI), an evidence-based communication technique to increase intrinsic motivation and self-efficacy for behavior change,\cite{miller2012motivational,miller2009toward} is one type of these interventions. MI sessions are generally aimed at eliciting ``change talk'', or statements of intrinsic motivation about patients' own desire, ability, reasons, need for and commitment to behavior change, which have been established by previous research\cite{apodaca2009mechanisms} as a reliable mediator of health behavior change. However, communication science approaches to understanding the efficacy of MI are inherently limited by traditional qualitative coding methods.

Qualitative coding of motivational interviews with pre-defined codes has been traditionally performed manually by trained annotators, which is a tedious and resource-intensive process that involves several iterations of reading, comprehension and interpretation of interview transcripts. Rapidly developing computational technologies, specifically, machine learning methods, offer a unique opportunity to accelerate this process. In particular, machine learning methods have been successfully applied to a variety of analytical tasks involving textual data, such as classification\cite{nigam2000text} and sentiment analysis.\cite{wang2012baselines} In our previous work, we examined the utility of machine learning methods for automated annotation \cite{hasan2016study,kotov2015interpretable} and analysis \cite{hasan2018predicting} of in-person MI sessions. Specifically, we demonstrated that machine learning methods can be utilized for annotation of MI transcripts according to a simple communication code scheme with the accuracy comparable to human coders.\cite{hasan2016study} Experimental data utilized in these studies, however, were prepared by transcribing audio conversations, which were clearly segmented into utterances by a counselor, a patient and a caregiver.

In this study, we focus on the analysis of e-Coaching sessions, behavior interventions that are delivered via email and grounded in the principles of motivational interviewing. Specifically, the e-Coaches involved in this study used emails to communicate motivation-enhancing messages that encourage healthy eating among GenY adolescents. E-coaching data is comprised of email responses, which are free-text documents, unlike more traditional dyadic clinical interviews that are naturally segmented into utterances due to their conversational nature.

The unstructured nature of e-Coaching exchanges poses a unique set of challenges for their qualitative analysis. Usually, email segments are automatically coded with the Minority Youth-Sequential Coding of Process Exchanges (MYSCOPE), \cite{carcone2013provider} a qualitative coding scheme to characterize patient-coach communication during email exchanges. A significant barrier to fully automating the behavior coding process of e-Coaching emails is their segmentation into textual fragments that correspond to distinct communication behaviors. Automating this task is a unique and challenging problem due to the following major reasons:

@@

The goal of this research study is to assess the applicability of machine learning methods for automated segmentation of e-Coaching emails into textual fragments corresponding to individual behaviors, which is the first step of the coding process of e-Coaching communications. In particular, we introduced word embedding or lexical, punctuation and  part-of-speech (POS) features and experimented with both traditional supervised machine learning method, linear-chain conditional random fields (CRF)\cite{lafferty2001conditional} and deep learning methods, such as multilayer perceptrons (MLP),\cite{rumelhart1986learning} bidirectional recurrent neural networks (BRNN)\cite{schuster1997bidirectional} and convolutional recurrent neural networks (CRNN),\cite{treviso2017sentence} to find the best performing method and feature combination.

Relevant previous work in the biomedical domain primarily focused on segmentation of text into sections and headers\cite{apostolova2009automatic,denny2009evaluation,tepper2012statistical,cho2002text} or sentence boundary detection\cite{griffis2016quantitative,kreuzthaler2015detection,treviso2017sentence,fraser2015sentence}. Apostolova et al. applied SVM along with word-vector cosine similarity metric combined with several heuristics to segment clinical reports into sections, such as demographics, history, procedure, finding and impression.\cite{apostolova2009automatic} After identification of each line in the document, Tepper et al. trained Maximum Entropy models for section classification.\cite{tepper2012statistical} Denny et al. proposed a SecTag algorithm, which combined natural language processing techniques, terminology-based rules and a Naive Bayes classifier to identify the sections and headers that achieved 99\% recall with 95.6\% precision. \cite{denny2009evaluation} On the other hand, SVM based on prosodic and part of speech features \cite{kreuzthaler2015detection} and recurrent convolutional neural networks using word embeddings \cite{griffis2016quantitative} were utilized for detecting sentence boundaries. Liu et al. have shown that a linear chain CRF performed better than the hidden markov model and Maxent on the NIST sentence boundary detection.\cite{liu2005using} Recently, Treviso et al. proposed a recurrent convolutional neural network for automatic identification of sentence boundary in a neuropsychological text in Portuguese language.\cite{treviso2017sentence} Their proposed model combined prosodic and part of speech features with pre-trained GloVe word vectors,\cite{pennington2014glove} which achieved 0.74 F1-score in sentence boundary detection on the Brazilian constitution dataset.

In this study, we adopt the model proposed by Treviso et al. \cite{treviso2017sentence} and rename it as convolutional recurrent neural networks to apply the model differently from the aforementioned studies. The main difference is that our study focuses on the segmentation of e-Coaching emails as distinct from determining the sentence boundary. Another important difference is that punctuation features were utilized in combination with word embeddings and part-of-speech features. Furthermore, we train our model with pre-trained word2vec word vectors,\cite{mikolov2013distributed} instead of using GloVe word vectors. Segmentation of e-Coaching emails is also different from a traditional shallow discourse analysis of conversations \cite{galley2003discourse} in that the focus is on segmentation, rather than on determining the types of transitions between the utterances or assigning utterances to speakers.

Recently, an online clinical intervention called MENU GenY (Making Effective Nutrition Choices for Generation Y) was proposed and evaluated. \cite{alexander2017motivations} MENU GenY is a technology-based public health intervention that relies on personalized e-coaching to encourage increased fruit and vegetable intake among young adults, aged 21-30. The goal of MENU GenY was to develop a better coding dictionary among GenY to improve eating habits. However, segmentation of clinical exchange in the context of electronically delivered interventions, in particular, segmentation of clinical interaction text into groups of MI behaviors, is still performed manually, which slows down qualitative analysis of these interventions. This study proposes several machine learning methods to address this problem and the authors are unaware of any other work that focused on the same problem.

@@

The experimental dataset for this work was constructed from 49 e-coaching sessions, which include a total of 3,138 segmented and annotated MI behaviors. Each session represents an MI intervention delivered via email. We formulate the segmentation task as a binary classification problem where each word or punctuation mark annotated with a label 1 or 0, to indicate whether it precedes a new segment or not. In total, we obtained 95,777 words and 7,140 punctuation marks, which include 3,138 ``new segment'' and 99,779 ``same segment'' instances. In this study, we experimented with traditional machine learning method, CRF and deep learning methods, such as MLP, BRNN and CRNN. For MLP models, samples were created with $2l$ words or punctuations at each step in a word sequence, where a sample contains next $l$ words or punctuations and prior $l$ words or punctuations including current wordor punctuation. Each sample is classified into either ``new segment'' or ``same segment'' based on whether the current word or punctuation precedes a new segment or not. For CRF, BRNN and CRNN models, an email was taken as input sequence, such that POS tags and word embedding of each word or punctuation mark were used as input and binary labels (1 or 0) corresponding to ``new segment'' and ``same segment'' classification decision were considered as the output of the model at each step. In the gold standard, words or punctuations within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1.

@@

We utilized three types of features in conjunction with CRF, MLP, BRNN and CRNN methods: word embeddings or lexical features, punctuation and POS features. Since POS tags have been shown to be effective semantic abstractions of individual words, we used POS features for our experiment.\cite{liu2005using,treviso2017sentence} To extract our POS features, we tag the e-Coaching emails using the NLTK POS tagger. \cite{bird2009natural} Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;'\} between a pair of words, are also employed as a feature since punctuation marks designate the boundary of a sentence, clause and phrase and often also correspond to a segment boundary. For natural language processing (NLP) tasks, inputs are received as a text where words can be considered as the basic processing unit. Therefore, it is important to represent a word in such a way that it carries all relevant information. Word embeddings are one of this representation where each word represented as a real-valued vector in a high dimensional vector space. The distributed representation of word allows capturing semantic, syntactic and morphological information from large unannotated corpora.\cite{pennington2014glove, mikolov2013distributed} Since pre-trained word2vec word vectors provide the best results, in our experiment, we utilized pre-trained word2vec word vectors for words. Trained word vectors were also utilized when words or punctuation marks are not available in the pre-trained word2vec word vectors. Figure~\ref{fig:embedding-dimension} illustrates the performance of CRNN model on e-Coaching email segmentation by varying the dimension of pre-trained and trained word embeddings with GloVe and word2vec models. It was observed that best performance achieved with pre-trained 300-dimensional word2vec word vectors when three types of features are used together. Therefore, we report our results with word2vec 300-dimensional word vectors for all deep learning models used in this study. CRF utilized lexical features which represent each word in an e-Coaching email.

@@

In the present study, we experimented with four different classifiers, including one traditional machine learning method, CRF and three deep learning models. Since deep learning methods provide a flexible architecture for constructing and synthesizing complex models, we take the advantage of this flexibility to test MLP, BRNN and CRNN models for the task of segmentation of e-Coaching emails.

\textbf{Conditional Random Fields}: CRF has been widely used in various NLP tasks including part-of-speech tagging and segmentation tasks.\cite{lafferty2001conditional, hirohata2008identifying} Unlike a maximum entropy Markov model which uses per-state exponential models for the conditional probabilities of next states given the current state, a CRF is a discriminative model of the conditional distribution of the entire sequence of labels given the observation sequence. A traditional linear-chain CRF model is defined as a conditional probability distribution $p(y|x)$ for output and input sequences, $y$ and $x$, by the Eq. \ref{Eq:crf}

@@

Here $Z_x$ is a normalization factor over all possible labelings of $x$, and $f_k(y_{t-1}, y_t, x, t)$ is a feature function, and $\lambda_k$ is a learned weight associated with feature $f_k$. The optimal output sequence $y^*$ for an input sequence $x$, $y^* = {arg\,max}_y p(y|x)$, is obtained efficiently by the Viterbi algorithm. In our experiments, the following features were utilized with CRF models: i) current word or punctuation ii) next and previous 3 words or punctuations iii) whether the word or punctuation is a special character (';', '?', '.', ',', '!', ':', etc) or not iv) whether the word is a title word or not? (e.g. ``The'' is a title word but ``the'' is not) v) POS tags

\textbf{Multilayer Perceptrons}: MLP is a feed-forward artificial neural network which maps set of input onto a set of appropriate output.\cite{rumelhart1986learning} This study utilized MLP with a nonlinear activation function (relu) and one hidden layer consisting of 128 hidden units. In order to prevent over-fitting, we implement dropout on fully connected layers, which randomly hide some neurons during the training phase. \cite{srivastava2014dropout} Dropout is also applied to the fully connected layers in BRNN and CRNN models. For a given sample with $2l$ words or punctuations, the first layer input of MLP was created by the summation of first $l$ word vectors and last $l$ word vectors. Figure~\ref{fig:length-mlp} demonstrates the performance of MLP model on e-Coaching email segmentation by varying value of $l$. Based on the results in Figure~\ref{fig:length-mlp}, MLP results were reported with $l = 2$, since best performance of MLP models were observed when $l = 2$.

@@

\textbf{Bidirectional Recurrent Neural Networks}: BRNN is a neural network designed to capture sequential patterns by considering both past \& future inputs as well as complex relationships between input features and output labels.\cite{schuster1997bidirectional} The output of BRNN layer is computed with the summation of the forward RNN output and backward RNN output. Gated Recurrent Units (GRU)\cite{chung2014empirical} are utilized as RNNs that are capable of handling variable size input sequence and have an internal memory that can be reset. Figure~\ref{fig:crnn} will represent an architecture of BRNN if convolution layer is removed.

\textbf{Convolutional Recurrent Neural Networks}: CRNN is a deep neural networks architecture,\cite{treviso2017sentence} shown in Figure~\ref{fig:crnn}, consisting of 5 layers: 1) input layer 2) embedding layer 3) convolution layer with max pooling 4) BRNN layer 5) fully connected layer with dropout and sigmoid output.

@@

E-coaching email exchanges, represented as a sequence of $m$ words and punctuations are fed into the input layer which provides an $m \times n_e$ matrix after fetching the pre-trained word vectors. When POS tags are used in combination with word embeddings, 10-dimensional POS vectors are concatenated with 300-dimensional word vectors to obtain new vectors $n_e = n_w + n_p$ of size 310. The 1D convolutional layer generates $m - h_c + 1$ new features with $h_{c.n_e}$, a filter of $h_c$ words and zero-padding on both sides of the text. A pooling operation was performed over time to find the most significant features in a region. After applying $n_f=100$ different filters on convolutional layer, $m \times n_f$ data are fed into bidirectional recurrent neural networks, which contain 200 hidden units. Then, dropout is used to randomly drop 50\% of the neurons and passes them through a fully connected layer, where a sigmoid operation is performed, provide the probability of whether the word or punctuation precedes a new segment or not. We experimentally determined that the best performance is achieved with 5 folds cross-validation when filter length in convolution layer is 7, pool size is 3, batch size is 32, adam\cite{kingma2014adam} is used for optimization and the early stopping strategy is used with dropout to prevent over-fitting. To utilize our source codes by other research studies, source codes of all experiments will be publicly available at https://github.com/teanalab/xxxxx upon acceptance of this paper for publication.

\subsection*{\textit{Evaluation metrics}}
We report standard metrics for experiments (precision, recall and F1-measure) to evaluate the performance of binary classifiers.\cite{aas1999text} However, accuracy is not reported as a performance metric because accuracy is highly sensitive to the prior class probabilities and does not fully describe the actual difficulty of the decision problem for an unbalanced dataset. The results are reported based on 5 folds cross-validation and weighted macro-averaging over the folds. We also estimate the area under the receiving operating characteristics (ROC) curve\cite{kumar2011receiver} (AUC) metric due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets. \cite{hu2015kernelized}

@@

\begin{table}[ht]
\centering
\caption{Performance of CRF, MLP, BRNN and CRNN methods for identification of ``new segment'' class as well as weighted average over ``new segment'' and ``same segment'' classes when word embeddings or lexical features are used. The highest value for each performance metric
is highlighted in bold.}
\label{tab:result_base}
  \begin{tabular}{|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{New Segment}} & \multicolumn{3}{|c|}{\textbf{Weighted Macro Average}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} \\ \hline    
 CRF & 0.782 & 0.691 & 0.733 & 0.983 & 0.984 & 0.984 \\ \hline
 MLP & \textbf{0.836} & 0.593 & 0.694 & 0.982 & 0.983 & 0.982 \\ \hline
 BRNN & 0.606 & 0.680 & 0.641 & 0.977 & 0.976 & 0.976 \\ \hline
 CRNN & 0.775 & \textbf{0.797} & \textbf{0.785} & \textbf{0.986} & \textbf{0.986} & \textbf{0.986} \\ \hline
  \end{tabular}
\end{table}

@@

\begin{table}[ht]
\centering
\caption{Performance of CRF, MLP, BRNN and CRNN methods for identification of ``new segment'' class as well as weighted average over ``new segment'' and ``same segment'' classes when all features are used together. The highest value for each performance metric
is highlighted in bold.}
\label{tab:result_weighted_avg}
 \begin{tabular}{|l|l|l|l|l|l|l|}
  \hline
   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{New Segment}} & \multicolumn{3}{|c|}{\textbf{Weighted Macro Average}} \\\cline{2-7}
   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Measure} \\ \hline    
 CRF & 0.813 & 0.772 & 0.792 & 0.988 & 0.988 & 0.988 \\ \hline
 MLP & \textbf{0.814} & 0.732 & 0.771 & 0.987 & 0.987 & 0.987 \\ \hline
 BRNN & 0.683 & 0.820 & 0.745 & 0.985 & 0.983 & 0.984 \\ \hline
 CRNN & 0.789 & \textbf{0.864} & \textbf{0.825} & \textbf{0.990} & \textbf{0.989} & \textbf{0.989} \\ \hline
  \end{tabular}
\end{table}
